Practical Ethics for Computational Semantics\\ \large Philosophy of Distributional Semantics

### 2017-06-05 pds/Paper/paper_pds.tex
approaching the end

\documentclass{article}
\title{Practical Ethics for Computational Semantics\\
\large Philosophy of Distributional Semantics}
\date{}
\author{Silvan Hungerb{\"u}hler}
\usepackage{enumitem}
\usepackage{verbatim}
\usepackage{hyperref}
\usepackage{comment}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{biblatex}
\addbibresource{pds_references.bib}
\usepackage{csquotes}
\begin{document}
\maketitle
\begin{abstract}
The increase in availability of data and advances in statistical methods has led to the use of Machine Learning (ML) technology in an immense variety of societal domains. ML algorithms are used to inform decisions as diverse and life-changing as access to financial credit, incarceration, university admission, medical treatment or the legal boundaries of espionage by the NSA.
The use of this technology is not always unproblematic, as its careless application can cause serious harm, even if such effects occur unintentionally.
The seriousness of the potential to disrespect people's privacy and integrity or systematically disadvantage specific groups has led to a recent debate concerning ethical guidelines and best practices for the general use of ML.
Little has been said, however, about the proper use of algorithms in the specific field of Distributional Semantics (DS). Some researchers have called attention to possible dangers of using ML in DS, as it harbors the potential to reinforce harmful biases along dimensions of race and gender.
The present paper wants to point to one particular attempt to provide general-purpose guidelines for ML and draw on these existing insights to suggest guidelines and best practices for computational methods in DS.
As the domain of DS differs in important aspects from other applications of ML, notably also natural language processing,
this involves assessing  the applicability of general purpose ML guidelines to DS with these difference in mind.
\end{abstract}
\section{Introduction}
As vast amounts of digital data have become available over past decades, algorithms are employed to use the data in order to take and inform decisions all across society. 
Algorithms have a say in determining who is granted mortgages, who is released on bail or invited for job applications \cite{algorithms2016}.
The theory behind many of these algorithms is \emph{machine learning} (ML).
Although most inventors and designers of ML algorithms certainly do not harbor malicious intent, the risk of unwillingly harming people is real.

There are several different sources of harmful effects that these techniques can bring about. With the very large amounts of data collected, stored and processed, issues of privacy arise naturally. Especially when sophisticated data gathering techniques are used to scrape the web, peoples's explicit consent to provide their data can easily go by the board. Another recurrent problem is the two-sided nature of any technology: it can be used for the good or for dubious purposes, as fraud or spying, for example. Finally, big data and ML approaches harbor the risk of re-enforcing structures of discrimination along racial, gender or others social lines. In a widely publicized case, to name one concrete example, automated systems have been used for legal decisions which lead to consistently biased outcomes against Blacks and Latinos in the USA \cite{angwin2016machine}.

Many of these issues are of great importance to modern societies. Consequently, people working at the intersection of information technology and ethics have attempted to lay down foundations on how to handle these technologies. The realization that algorithms fed on big data come with a host of problematic issues has led increased public as well as institutional scrutiny (notably, the oval office \cite{united2014big}) and, consequently, to the elaboration of a variety of principles and concrete guidelines for their use. Many of these efforts have naturally been directed at the broad topic of ML in general \cite{leidner2017ethical}. Only a few years ago the discussion has turned on the specific use of ML for analysis and modeling of language - sometimes broadly referred to as \emph{natural language processing} (NLP).

@@

\emph{Natural language processing} is a field of research that uses computational tools to build natural language text or speech applications. It is related to computational linguistic, an area of linguistic that uses computational techniques to get a scientific understanding of natural language. The knowledge thus obtained can be used  to develop computer systems that serve a wide variety of purposes. Principal applications include translation between natural languages, artificial intelligence, user interfaces, like chat-bots, for example, summarization of text or speech recognition \cite{chowdhury2003natural}. The area of NLP frequently serves itself of methods from ML, for example to process very large textual corpora. Indeed, parts of NLP can be thought of as particular applications of ML techniques to the domain of natural language.

@@

Following the distributional hypothesis, the meaning of an expression is a function of the context in which it occurs, where context can be construed in different ways. Most typically, it is other words surrounding an expression in a text - the sentence in which it appears, for example. Context is not restricted to such simple analysis, however, as it can possibly contain more complex syntactic relations within a text, or even visual, textual or auditory information gathered from data. \textbf{Moar refs?}

@@

DS is concerned with the specific topic of meaning in the study of natural language. If viewed from an application-oriented standpoint it could be seen as one of many sub-fields of NLP research. Indeed, as far as the present paper is concerned, it is precisely the application of DS in industry that requires heightened attention.

\section{Principles for  for Machine Learning}\hypertarget{sec3}{}
Algorithmically informed decisions are used in ever more domains of the private and public sectors and thus have the potential to impact society in significant ways. We now specify four ways in which unreflected usage of ML can bring about harm.

A first concern is the issue of privacy. Since very large amounts of data are required for these methods to be effective, great efforts are devoted to collecting large data sets which may contain highly sensitive information about individuals or organizations. It is thus of importance who gets to see these data and how they are protected from third parties accessing them. There have been reported cases of researchers working with large corpora of texts containing peoples' complete medical record who sent these files indiscriminately around by email. A second aspect of privacy is how the data is collected in the first place. It is a question of ethical concern whether the individuals whose data is collected agree to their personal information being used.

A second point concerns the disruptive potential ML techniques have on labor markets. As ever more cognitive tasks can be automated, the economic landscape of service sector economies is bound to undergo big change. Some workers will without a doubt be disadvantaged by these changes.

A third difficult point is the so-called \emph{double use} problem. New technology's use is often underspecified and could be used for benefit of humanity or to its detriment. Chat-bots developed by researchers in NLP, to name a concrete example, can be used to assist people to access governmental health-care services, yet they have reportedly been used by the Mexican government to disrupt political discussions deemed subversive on social media \cite{leidner2017ethical}.

Finally, ML harbors the potential to amplify structural discrimination and systematically disadvantage people because of their race, gender or other social category. If the models which are trained algorithmically are fed biased data, then predictions or classifications they make will reflect that bias.
Data processed by ML techniques can reflect societal prejudice and bias, so that unquestioningly using them for decision making can reinforce society's existing discriminatory structure. A related issue is that data available to the algorithms may systematically omit information relevant to what it is trying to uncover, as when minority groups are not adequately represented in the data. Yet another issue is that ML techniques may use useful patterns and relationships in the data that are really mirroring historical factors of discrimination and marginalization \cite{barocas2016big}.

In the broader field of ML some people seem to have recognized the importance ethical considerations. For the domain-general use of ML techniques there exists a series of studies by from legal, political computer science experts which specify principles and guidelines (cf. \cite{pasquale2015black}\cite{feldman2015certifying}\cite{barocas2016big}. Starting in 2014 with a workshop series at the \emph{Conference on Neural Information Processing Systems}, an organization called \emph{Fairness, Accountability, and Transparency in Machine Learning} (FATML) has been working on solutions for ethical problems in ML. Furthermore, many non-technical articles have been published outside of academic channels.

In the case of NLP there has been  has seen some movement in the last couple of years. There have been discussion regarding the moral agency of chat-bots \cite{thieltges2016devil}, as well as more general contributions to ethical guidelines in broader NLP \cite{hovy2016social}.
Additionally, both for ML and NLP many public events were hosted and conferences established to discuss ethical issues (cf. \cite{conferences}).
As a field, however, DS in particular has seen little attention in this respect. Although there are already proposals for best practices in NLP \cite{leidner2017ethical}, for example, no similar effort has been made for DS.

@@

We will now briefly go over the five principles - \emph{responsibility, explainability, accuracy, auditability and fairness} to describe their original intention in the PAA.
Responsibility requires developers to take seriously the possibility that adverse effects are produced their algorithmic system. Consequently, there should be designated roles within the organization to be responsible for redress if harm is caused. Excuses akin to ''the algorithm did it'' are no valid scapegoats. This reflects the acknowledgment that, at some level, there are people who design these systems and put them to use.
Explainability demands that both data selection and decisions taken by the algorithms can be explained to people affected by its usage in terms that are understandable by laywomen. The point here is not to obscure the workings of the systems to a point where no stakeholder can grasp it.
Accuracy asks designers and users of the system to keep track of its mistakes and sources of error, so that worst case scenarios become more easily anticipated and possible damage reduced.
Audibility is the requirement to create transparency in such a way that interested third parties are in the position to review the algorithms performance. This point is related to explainability.
Finally, Fairness requires that the algorithmic system does not affect different demographic groups in disparate and unfavorable ways, that is, discriminates against them.
We now want to assess to what degree we can use these principles for the field of DS.
\section{Principles for Distributional Semantics}\hypertarget{sec4}{}
It should be fair to say that some of these principles allow seamless transfer from general ML to any specific sub-domain, as DS, without much further discussion.
This goes for the principles corresponding to accuracy - logging mistakes is good practice no matter where -, auditability - one should never obstruct uninterested parties' inquiries into the possibility of things going wrong - and explainability - obfuscating one's work and keeping affected parties ignorant is evidently bad practice both in ML and in DS.
Much more interesting than the overlap lately mentioned, however, is the question how the two remaining principles apply to to DS.

Concerning the question of fairness, consider a concrete scenario where the problem arises: A recent study has found, for example, that a model of lexical word meaning trained on Google News articles displays significant bias along binary gender dimensions \cite{bolukbasi2016man}. Despite the articles in the corpus being written by professional journalists, the word \emph{woman} bears the same semantic relation to \emph{homemaker} in the meaning space as the word \emph{man} to \emph{computer scientist}. A search engine could use semantic information available in such a model, with bias and all, to optimize the relevance of their search results. Imagine an employer querying Google for ''Computer Scientist Kabul'' to find professionals in the area and Google deems LinkedIn profiles with male-associated names more relevant to the search request. Consequently, male computer scientists will find their profiles ranked higher in the results of the query and thus obtain more visibility. This would be a clear-cut case of discrimination and disadvantage based on gender, brought about directly by recommendation mechanism which utilizes DS. These are not mere hypotheticals though, but already very concrete issues. It is has been reported that when one queries Google with the name of a person, then Google is more likely to show advertisement related to that person's criminal record if the name is associated with black people \cite{sweeney2013discrimination}.

@@

This story goes to show that it can be excruciatingly difficult to find the responsible human agent when DS is used for algorithmic systems. Even if one grants that the meanings as found in a corpus are biased in an unacceptable way, the question still remains what acceptability would entail. A follow to this is whether one can remove prejudice and bias from meaning spaces without rendering them unusable for usage in applications. What, for example, if the fact that the expression "woman" is functional in natural language discourse precisely because of its unfortunate semantic relationship to "man". A recent paper by Bolukbasi et al. contradicts this pessimistic view \cite{bolukbasi2016man}. According to it, it is possible to debias the meanings of a semantic space along binary gender dimensions while still preserving its useful properties.

@@

In this fashion fairness principle could be addressed in DS. This does not yet respond, however, to the question where exactly responsibility lies - that is, with whom - in case something does go wrong.
It is perhaps the most intricate difficulty concerning ethics in DS as potential culprits are society at large, the producers of the corpus, the distributional semanticist or the person who implements the system that interacts with society again. 
The nature of language meaning and how language and society interact are notoriously contested - the heated debates about what is appropriate in gendering language serves as a case in point. For data engineers and programmers who work with these systems there is certainly no easy answer to this issue at first sight.
The best that can be offered in terms of practical advice is that if the possibilty exists to algorithmically rid a DS model of biases relevant to its application, then this should be done.
\section{Conclusion}\hypertarget{sec5}{}
The preceding discussion seems to warrant but two clear take-aways for DS. First, if a semantic space does contain bias and will be further used in some application, then efforts should be undertaken to debias it. Second, if no the DS model cannot be debiased, then it should not be employed externally without further efforts to ensure that it does not discriminate. The responsibility for these two points ultimately lie with the designer of the application.
\section{References}\hypertarget{sec6}{ }
\printbibliography

>>>>

\documentclass{article}
\title{Practical Ethics for Computational Semantics\\
\large Philosophy of Distributional Semantics}
\date{}
\author{Silvan Hungerb{\"u}hler}
\usepackage{enumitem}
\usepackage{verbatim}
\usepackage{hyperref}
\usepackage{comment}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{biblatex}
\addbibresource{pds_references.bib}
\usepackage{csquotes}
\begin{document}
\maketitle
\begin{abstract}
The increase in availability of data and advances in statistical methods has led to the use of Machine Learning (ML) technology in an immense variety of societal domains. ML algorithms are used to inform decisions as diverse and life-changing as access to financial credit, incarceration, university admission, medical treatment or the legal boundaries of espionage by the NSA.
The use of this technology is not always unproblematic, as its careless application can cause serious harm, even if such effects occur unintentionally.
The seriousness of the potential to disrespect people's privacy and integrity or systematically disadvantage specific groups has led to a recent debate concerning ethical guidelines and best practices for the general use of ML.
Little has been said, however, about the proper use of algorithms in the specific field of Distributional Semantics (DS). Some researchers have called attention to possible dangers of using ML in DS, as it harbors the potential to reinforce harmful biases along dimensions of race and gender.
The present paper wants draw on one particular attempt to provide general-purpose guidelines for ML - the \emph{Principles for Accountable Algorithms} to suggest guidelines and best practices for computational methods in DS.
As the phenomena of interest in DS differs in important aspects from other applications of ML, notably also natural language processing,
this involves assessing the applicability of general purpose ML guidelines to DS. Our main finding is that most principles carry over smoothly, except for issues concerning responsibility and algorithmic fairness.
\end{abstract}
\section{Introduction}
As vast amounts of digital data have become available over past decades, algorithms are employed to use the data in order to take and inform decisions all across society. 
Algorithms have a say in determining who is granted mortgages, who is released on bail or invited for job applications \cite{algorithms2016}.
The theory behind many of these algorithms is \emph{machine learning} (ML).
Although most inventors and designers of ML algorithms certainly do not harbor malicious intent, the risk of unwillingly harming people is real.

There are several different sources of harmful effects that these techniques can bring about. %With the very large amounts of data collected, stored and processed, issues of privacy arise naturally. Especially when sophisticated data gathering techniques are used to scrape the web, peoples' explicit consent to provide their data can easily go by the board. Another recurrent problem is the two-sided nature of any technology: it can be used for the good or for dubious purposes, as fraud or spying, for example. Finally, big data and ML approaches harbor the risk of re-enforcing structures of discrimination along racial, gender or others social lines. %In a widely publicized case, to name one concrete example, automated systems have been used for legal decisions which lead to consistently biased outcomes against Blacks and Latinos in the USA \cite{angwin2016machine}.
Problems include privacy concerns, abuse of technology or unintended structural discrimination; all are of great importance to modern societies. Consequently, people working at the intersection of information technology and ethics have attempted to lay down foundations on how to handle these technologies. The realization that algorithms fed on big data come with a host of problematic issues has led increased public as well as institutional scrutiny (notably, the oval office \cite{united2014big}) and, consequently, to the elaboration of a variety of principles and concrete guidelines for their use. Many of these efforts have naturally been directed at the broad topic of ML in general \cite{leidner2017ethical}. Only a few years ago the discussion has turned on the specific use of ML for analysis and modeling of language - sometimes broadly referred to as \emph{natural language processing} (NLP).

@@

\emph{Natural language processing} is a field of research that uses computational tools to build natural language text or speech applications. It is related to computational linguistic, an area of linguistic that uses computational techniques to get a scientific understanding of natural language. The knowledge thus obtained can be used  to develop computer systems that serve a wide variety of purposes. Principal applications include translation between natural languages, artificial intelligence, user interfaces, like chat-bots, for example, summarization of text or speech recognition \cite{chowdhury2003natural}. The area of NLP frequently serves itself of methods from ML, for example to process very large textual corpora. %Indeed, parts of NLP can be thought of as particular applications of ML techniques to the domain of natural language.

@@

Following the distributional hypothesis, the meaning of an expression is a function of the context in which it occurs, where context can be construed in different ways. Most typically, it is other words surrounding an expression in a text - the sentence in which it appears, for example. Context is not restricted to such simple analysis, however, as it can possibly contain more complex syntactic relations within a text, or even visual, textual or auditory information gathered from data \cite{erk2012vector}.

@@

%DS is concerned with the specific topic of meaning in the study of natural language. If viewed from an application-oriented standpoint it could be seen as one of many sub-fields of NLP research. % Indeed, as far as the present paper is concerned, it is precisely the application of DS in industry that requires heightened attention.
\section{Principles for Machine Learning}\hypertarget{sec3}{}
Algorithmically informed decisions are used in ever more domains of the private and public sectors and thus have the potential to impact society in significant ways. We now specify four ways in which unreflected usage of ML can bring about harm.

A first concern is the issue of privacy. Since effective use of these methods requires very large amounts of data, great efforts are devoted to collecting data sets which may contain highly sensitive information about individuals or organizations. It is thus of importance who gets to see these data and how they are protected from third parties accessing them. 
%There have been reported cases of researchers working with large corpora of texts containing peoples' complete medical record who sent these files indiscriminately around by email. 
Another aspect of privacy is how the data is collected in the first place. Principally, whether the individuals whose data is collected agree to their personal information being used.

A second point concerns the disruptive potential ML techniques have on labor markets. As ever more cognitive tasks can be automated, the economic landscape of service sector economies is bound to undergo big change \cite{west2015happens}. Some workers will without a doubt be disadvantaged by these changes.

A third difficulty is the so-called \emph{double use} problem. New technology's use is often underspecified and could be used for benefit of humanity or to its detriment. Chat-bots developed by researchers in NLP, to name a concrete example, can be used to assist people to access governmental health-care services, yet they have reportedly been used by the Mexican government to disrupt political discussions deemed subversive on social media \cite{leidner2017ethical}.

Finally, ML harbors the potential to amplify structural discrimination and systematically disadvantage people because of their race, gender or other social category. If the models which are trained algorithmically are fed biased data, then predictions or classifications they make will reflect that bias.
Data processed by ML techniques can reflect societal prejudice and bias, so that unquestioningly using them for decision making can reinforce society's existing discriminatory structure. A related issue is that data available to the algorithms may systematically omit information relevant to what it is trying to uncover, as when minority groups are not adequately represented in the data. Yet another issue is that ML techniques may exploit useful patterns and relationships in the data that are really mirroring historical factors of discrimination and marginalization \cite{barocas2016big}.

In the broader field of ML some people seem to have recognized the importance of ethical considerations. For the domain-general use of ML techniques there exists a series of studies by legal, political and ethical scholars, as well as computer science experts which specify principles and guidelines (cf. \cite{pasquale2015black}\cite{feldman2015certifying}\cite{barocas2016big}. Starting in 2014 with a workshop series at the \emph{Conference on Neural Information Processing Systems}, the organization FATML has been working on solutions for ethical problems in ML. Furthermore, many non-technical articles have been published outside of academic channels (cf. \cite{new2016}\cite{algorithms2016}).

In the case of NLP there has been some movement in the last couple of years with discussions regarding the moral agency of chat-bots \cite{thieltges2016devil}, as well as more general contributions to ethical guidelines in broader NLP \cite{hovy2016social}.
Additionally, both for ML and NLP many public events were hosted and conferences established to discuss ethical issues (cf. \cite{conferences}).
As a field, however, DS in particular has seen little attention in this respect. Although there are already proposals for best practices in NLP \cite{leidner2017ethical}, for example, no similar effort has been made for DS.

@@

We will now briefly go over the five principles - \emph{responsibility, explainability, accuracy, auditability} and \emph{fairness} to describe their original intention in the PAA.
Responsibility requires developers to take seriously the possibility that adverse effects are produced their algorithmic system. Consequently, there should be designated roles within the organization to be responsible for redress if harm is caused. Excuses akin to ''the algorithm did it'' are no valid scapegoats. This reflects the acknowledgment that, at some level, there are people who design these systems and put them to use.
Explainability demands that both data selection and decisions taken by the algorithms can be explained to people affected by its usage in terms that are understandable by laywomen. The point here is not to obscure the workings of the systems to a point where no stakeholder can grasp it.
Accuracy asks designers and users of the system to keep track of its mistakes and sources of error, so that worst case scenarios become more easily anticipated and possible damage reduced.
Audibility is the requirement to create transparency in such a way that interested third parties are in the position to review the algorithms performance. This point is related to explainability.
Finally, Fairness requires that the algorithmic system does not affect different demographic groups in disparate and unfavorable ways, that is, discriminates against them.
We now want to assess to what degree we can use these principles for the field of DS.
\section{Principles for Distributional Semantics}\hypertarget{sec4}{}
It should be fair to say that some of these principles allow seamless transfer from general ML to any specific sub-domain, as DS, without much further discussion.
This goes for the principles corresponding to accuracy - logging mistakes is good practice no matter where -, auditability - one should never obstruct uninterested parties' inquiries into the possibility of things going wrong - and explainability - obfuscating one's work and keeping affected parties ignorant is evidently bad practice both in ML and in DS.
Much more interesting than the overlap lately mentioned, however, is the question how the two remaining principles apply to to DS.

Concerning the question of fairness, consider a concrete scenario where the problem arises: A recent study has found, for example, that a model of lexical word meaning trained on Google News articles displays significant bias along binary gender dimensions \cite{bolukbasi2016man}. Despite the articles in the corpus being written by professional journalists, the word \emph{woman} bears the same semantic relation to \emph{homemaker} in the meaning space as the word \emph{man} to \emph{computer scientist}. A search engine could use semantic information available in such a model, with bias and all, to optimize the relevance of their search results. Imagine an employer querying Google for ''Computer Scientist Kabul'' to find professionals in the area and Google deems LinkedIn profiles with male-associated names more relevant to the search request. Consequently, male computer scientists' profiles are ranked higher in the results and thus more visible. This would be a clear-cut case of discrimination and disadvantage based on gender, brought about directly by recommendation mechanism which utilizes DS. These are not mere hypotheticals though, but already very concrete issues. It is has been reported that when one queries Google with the name of a person, then Google is more likely to show advertisement related to that person's criminal record if the name is associated with black people \cite{sweeney2013discrimination}.

@@

This story goes to show that it can be excruciatingly difficult to find the responsible human agent when DS is used for algorithmic systems. Even if one grants that the meanings as found in a corpus are biased in an unacceptable way, the question still remains what acceptability would entail. A follow-up question to this is whether one can remove prejudice and bias from meaning spaces without rendering them unusable for usage in applications. What, for example, if the fact that the expression \emph{woman} is functional in natural language discourse precisely because of its unfortunate semantic relationship to \emph{man}. A recent paper by Bolukbasi et al. contradicts this pessimistic view \cite{bolukbasi2016man}. According to it, it is possible to debias the meanings of a semantic space along binary gender dimensions while still preserving its useful properties.

@@

In this fashion fairness principle could be addressed in DS. This does not yet respond, however, to the question where exactly responsibility lies - that is, with whom - in case something does go wrong.
It is perhaps the most intricate difficulty concerning ethics in DS as potential culprits are society at large, the producers of the corpus, the distributional semanticist or the person who implements the system that interacts with society again. 
The nature of language meaning and how language and society interact are notoriously contested - the heated debates about what is appropriate in gendering language serves as a case in point. For data engineers and programmers who work with these systems there is certainly no easy answer to this issue at first sight.
The best that can be offered in terms of practical advice is that if the possibility exists to algorithmically rid a DS model of biases relevant to its application, then this should be done.
\section{Conclusion}\hypertarget{sec5}{}
Aside from questions regarding fairness and responsibility we are confident that the PAA is an apt set of principles to guide DS. Private information, to name just one example, should be handled just as carefully by a distributional semanticist as by a medical doctor.

The preceding discussion seems to warrant heightened attention for the two remaining points in the PAA. First, if a semantic space does contain bias and will be further used in some application, then efforts should be undertaken to debias it. Second, if the DS model cannot be debiased, then it should not be employed externally without further efforts to ensure that it does not discriminate. The responsibility for these two points ultimately lies with the designer of the application. Excuses to the effect that the model just mirrors societal bias are unacceptable.

What remains to be done, then, is to constantly re-assess how DS systems are in fact affecting different demographic groups. As it is further investigated how DS representations of meaning and societal bias are related to each other, we might be able to elaborate more concrete tests and guidelines for the working distributional semanticist to check her own work for discriminatory effects. Perhaps it will also be possible to connect an ethics of DS more systematically with an ethics of NLP in general. Until then, the responsible person working on DS will have to assess the impact of her work on a case by case basis.

\section{References}\hypertarget{sec6}{ }
\printbibliography
